---
title: "Exponentially weighted moving average using Rcpp"
author: "Ryan Moerer\n"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Iterative operations like for loops have quite a bit of overhead in a language like R. Thus, it is good practice to resort to a vectorised way of doing things whenever possible. However, sometimes loop operations cannot be easily vectorized. Therefore, it might be better to implement your operation in a language like C++ where loops have a lot less overhead. You can easily do this using the R packaage `Rcpp`.

## Exponentially-Weighted Moving Average in R

Here's an R implementation of an exponentially-weighted moving average using a loop.

```{r}
ewmaR <- function(x, alpha) {
  ewma <- numeric(length(x))
  ewma[1] <- x[1]
  for (i in 2:length(x)) {
    ewma[i] <- alpha * x[i] + (1 - alpha) * ewma[i - 1]
  }
  ewma
}

ewmaR(1:20, 0.9)
```

## Exponentially-Weighted Moving Average using Rcpp

And here's a C++ implementation of that same operation using `Rcpp`.

```{r}
library(Rcpp)

cppFunction('NumericVector ewmaCPP(NumericVector x, double alpha) {
  NumericVector ewma(x.size());
  ewma[0] = x[0];
  for (int i = 1; i < x.size(); ++i) {
    ewma[i] = alpha * x[i] + (1 - alpha) * ewma[i - 1];
  }
  return ewma;
}')

ewmaCPP(1:20, 0.9)
```

## Comparing performance

As one can see, the EWMA implemented in C++ is magnitudes faster than the EWMA implemented in R.

```{r}
x <- runif(1e3)

bench::mark(
  ewmaR(x, 0.9),
  ewmaCPP(x, 0.9)
)
```
